Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-22 paper.qmd
Update paper.qmd

This call to rix() generates a `default.nix` file for a development shell that encapsulates 
a complete and reproducible research environment. It provides [R]{.proglang} and its packages 
(including [AER]{.pkg} at version 1.2-8), several TeXLive packages for \LaTeX document authoring, 
development versions of [rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
and the Positron editor. The reproducibility of this environment is guaranteed by pinning all 
components to a single point in time: the [R]{.proglang} packages are resolved from the CRAN 
snapshot of October 20, 2025, while all other system tools are fixed to the version of 
`nixpkgs` from that same date. Typing `nix-shell` in a terminal in the folder that containts
this `default.nix`, will drop the user into the shell. It should be noted that it is 
also possible to configure IDEs to dynamically load this shell to provide the right
tooling for developing.

@@

*   The RBC Model Simulation (`functions.jl`): This [Julia]{.proglang} script contains a single pure
  function that implements the state-space solution to the RBC model, based on @depaoli2009rbc.
  It takes the model's economic parameters as inputs and returns the simulated time-series
  data as a DataFrame.
*   The XGBoost Forecasting Model (`functions.py`): This [Python]{.proglang} script handles the 
  complete machine learning workflow through a series of modular functions. Its responsibilities include
  feature engineering (creating lagged variables), splitting the data into training and testing sets, 
  training the [XGBoost]{.pkg} model, and generating predictions.
*   The Visualization (`functions.R`): This [R]{.proglang} script contains a function that uses
  [ggplot2]{.pkg} to create the final visualization. It is designed to take the data frame of 
  actual and predicted values produced by the Python script and generate a plot comparing the two series.

@@

1.  It begins by defining the RBC model parameters in [Julia]{.proglang}.
2.  It specifies that the **RBC simulation** in `functions.jl` depends on these parameters.
3.  It then defines the series of **Python steps** (feature preparation, training, prediction), 
  making each one dependent on the output of the previous one. The first Python step is explicitly 
  made dependent on the output from the Julia simulation. [rixpress]{.pkg} handles the passing of data 
  between languages, in this case using the Arrow file format for efficiency.
4.  Finally, it defines the **R visualization** step, which depends on the final predictions from 
  the Python model, and a **Quarto rendering** step that depends on the generated plot.

When this script is run in the development shell (by executing `source("gen-pipeline.R")`), 
[rixpress]{.pkg} translates the declared pipeline into a master Nix expression
that Nix can execute, automatically handling the caching of results and re-running only the necessary 
steps if a piece of code or data changes.

@@

To execute this entire example, a user with Nix installed can simply run the 
`run_polyglot_example.sh` script available in the root of the paper's repository.
This script automates the full process:

1.  It first executes `gen-env.R` inside a temporary Nix shell to build the `default.nix` file.
2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers Nix to run the
  entire polyglot pipeline in the correct order.

Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` 
report containing the visualization.

>>>>

This call to `rix()` generates a `default.nix` file for a development shell that encapsulates 
a complete and reproducible research environment. It provides [R]{.proglang} and its packages 
(including [AER]{.pkg} at version 1.2-8), several TeXLive packages for \LaTeX document authoring, 
development versions of [rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
and the Positron editor. The reproducibility of this environment is guaranteed by pinning all 
components to a single point in time: the [R]{.proglang} packages are resolved from the CRAN 
snapshot of October 20, 2025, while all other system tools are fixed to the version of 
`nixpkgs` from that same date. Typing `nix-shell` in a terminal in the folder that containts
this `default.nix`, will drop the user into the shell. It should be noted that it is 
also possible to configure IDEs to dynamically load this shell to provide the right
tooling for developing.

@@

*   The RBC Model Simulation (`functions.jl`): This [Julia]{.proglang} script contains a single pure
  function that implements the state-space solution to the RBC model, based on @depaoli2009rbc.
  It takes the model's economic parameters as inputs and returns the simulated time-series
  data as a [DataFrame]{.pkg}.
*   The [XGBoost]{.pkg} Forecasting Model (`functions.py`): This [Python]{.proglang} script handles the 
  complete machine learning workflow through a series of modular functions. Its responsibilities include
  feature engineering (creating lagged variables), splitting the data into training and testing sets, 
  training the [XGBoost]{.pkg} model, and generating predictions.
*   The Visualization (`functions.R`): This [R]{.proglang} script contains a function that uses
  [ggplot2]{.pkg} to create the final visualization. It is designed to take the data frame of 
  actual and predicted values produced by the Python script and generate a plot comparing the two series.

@@

1.  It begins by defining the RBC model parameters in [Julia]{.proglang}.
2.  It specifies that the RBC simulation in `functions.jl` depends on these parameters.
3.  It then defines the series of [Python]{.proglang} steps (feature preparation, training, prediction), 
  making each one dependent on the output of the previous one. The first [Python]{.proglang} step is explicitly 
  made dependent on the output from the [Julia]{.proglang} simulation. [rixpress]{.pkg} handles the passing of data 
  between languages, in this case using the [Arrow]{.pkg} file format for efficiency.
4.  Finally, it defines the **R visualization** step, which depends on the final predictions from 
  the [Python]{.proglang} model, and a **Quarto rendering** step that depends on the generated plot.

When this script is run in the development shell (by executing `source("gen-pipeline.R")`), 
[rixpress]{.pkg} translates the declared pipeline into a master [Nix]{.proglang} expression
that [Nix]{.pkg} can execute, automatically handling the caching of results and re-running only the necessary 
steps if a piece of code or data changes.

@@

This entire example can be executed by running the `run_polyglot_example.sh` 
script available in the root of the paper's repository.
This script automates the full process:

1.  It first executes `gen-env.R` inside a temporary [Nix]{.pkg} shell to build the `default.nix` file.
2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers [Nix]{.pkg} to run the
  entire polyglot pipeline in the correct order.

Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` report found in
the `pipeline-output` folder containing the visualization.
